Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-08-23 jso.tex
adding sec6 to intro paragraph

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index --~smoothness and
squintability~-- that capture the complexities inherent in PP
optimisation problems. These two metrics are evaluated along with JSO
hyper-parameters to determine their effects on JSO success rate. Our
numerical results confirm the positive impact of these metrics on the
JSO success rate, with squintability being the most significant. The JSO
algorithm has been implemented in the \texttt{tourr} package and
functions to calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes, smoothness and squintability.
Section~\ref{sec-sim-deets} details two simulation experiments to assess
JSO's performance: one comparing JSO's performance improvements relative
to an existing optimiser, Creeping Random Search (CRS), and the other
studying the impact of different PP index properties on optimisation
performance. Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, rightrule=.15mm, leftrule=.75mm, arc=.35mm, titlerule=0mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, opacityback=0, coltitle=black, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, toprule=.15mm, left=2mm]

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index --~smoothness and
squintability~-- that capture the complexities inherent in PP
optimisation problems. These two metrics are evaluated along with JSO
hyper-parameters to determine their effects on JSO success rate. Our
numerical results confirm the positive impact of these metrics on the
JSO success rate, with squintability being the most significant. The JSO
algorithm has been implemented in the \texttt{tourr} package and
functions to calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, sharp corners, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes, smoothness and squintability.
Section~\ref{sec-sim-deets} details two simulation experiments to assess
JSO's performance: one comparing JSO's performance improvements relative
to an existing optimiser, Creeping Random Search (CRS), and the other
studying the impact of different PP index properties on optimisation
performance. Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-discussion} discusses implementation details and
insights for practitioners. Section~\ref{sec-conclusion} summarises the
work and provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, breakable, colframe=quarto-callout-note-color-frame, arc=.35mm, titlerule=0mm, left=2mm, bottomrule=.15mm, toprule=.15mm, coltitle=black, bottomtitle=1mm, opacitybacktitle=0.6, colback=white, toptitle=1mm, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white]
